Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-24 paper/paper.tex
Another comment
Answer more questions

\section{Discussion}
\label{sec:discuss}
\yl{I'm not sure how strong of a statement we can make about our proof
  technique. We've already demonstrated how to address type-level dependency
  when defining a logical relation through a simple example. Claiming that our
  proof structure is better seems quite ambitious, but I think there's a
  middle point where we claim that adding moderate features like typed
  reduction doesn't instantly make our code size expand all the way from 1000
  to 20,000 without saying the other developments are just verbose for no good
  reason} 
\scw{I think we can find reasons for much of the differences. Am I missing any?
While it would be difficult to assign numbers to each of the deltas, I think it 
is believable that when put together they add up to a lot.
\begin{itemize}
\item We don't include inductive or coinductive datatypes. We don't include 
cumulativity. We don't include Prop. We don't include universe polymorphism.
\item We state our equality algorithmically instead of declaratively. On one
  hand, this gives us automatic inversion principles when working with
  definition. Furthermore, we don't need to prove the equivalence between an
  algorithmic version and a declarative specification.
\item Our equality requires a simple decision algorithm and isn't type directed.
\item We don't prove decidability of type checking. (And, it is not provable
  for our system, because we lack type annotations on functions. We should
  point this out.)
\item Our logical relation is unary and untyped. The latter means that we don't 
  require the bookkeeping of a Kripke logical relation when reasoning about 
  open terms. I don't know why unary relations are shorter.
\item CoqHammer leads to short proofs.
\end{itemize}
}

@@

Finally, despite our earlier claim that our metatheoretic properties
are not as strong as some of the related work, we can strengthen our
results through syntactic means. For example, the
normalize-and-compare strategy does not induce an efficient algorithm
for type conversion, like the ones from \citet{decagda} and
\citet{martin-lof-a-la-coq}. However, with the standard syntactic
techniques from \citet{takahashi-parallel-reduction,
factorization-essentially}, we can prove that leftmost-outermost
reduction is a normalizing reduction strategy, from which we can
separately show the correctness a more efficient algorithm that
reduces the two terms to weak head normal form before recursively
comparing their subcomponents. We believe factoring such properties
out of a logical relation is valuable, as it helps us identify the
part of our proof that requires extra strength from the metatheory.
\scw{Why don't we just use leftmost-outermost reduction in the first place?
Do we even need nondeterministic parallel reduction?}

>>>>

\section{Discussion}
\label{sec:discuss}
\yl{I'm not sure how strong of a statement we can make about our proof
  technique. We've already demonstrated how to address type-level dependency
  when defining a logical relation through a simple example. Claiming that our
  proof structure is better seems quite ambitious, but I think there's a
  middle point where we claim that adding moderate features like typed
  reduction doesn't instantly make our code size expand all the way from 1000
  to 20,000 without saying the other developments are just verbose for no good
  reason}
\scw{I think we can find reasons for much of the differences. Am I missing any?
While it would be difficult to assign numbers to each of the deltas, I think it
is believable that when put together they add up to a lot.
\begin{itemize}
\item We don't include inductive or coinductive datatypes. We don't include
cumulativity. We don't include Prop. We don't include universe polymorphism.
\item We state our equality algorithmically instead of declaratively. On one
  hand, this gives us automatic inversion principles when working with
  definition. Furthermore, we don't need to prove the equivalence between an
  algorithmic version and a declarative specification.
\item Our equality requires a simple decision algorithm and isn't type directed.
\item We don't prove decidability of type checking. (And, it is not provable
  for our system, because we lack type annotations on functions. We should
  point this out.)
\item Our logical relation is unary and untyped. The latter means that we don't
  require the bookkeeping of a Kripke logical relation when reasoning about
  open terms. I don't know why unary relations are shorter.
\item CoqHammer leads to short proofs.
\end{itemize}
\yl{
  Just one more technical point to add, though it's in the text already:
  the logical relation is closed backward by full reduction rather
  than weak-head/deterministic reduction. This requires an early
  confluence result to show that the logrel is
  deterministic/functional but simplifies everything else
  (e.g. conversion is justified immediately by our preservation
  theorem, but that is not the case if you use weak head reduction).

  Also, regarding the first point, cumulativity only exists in
  Barras's work. Inductive, (maybe coinductive?), can be found in
  nuprl, metacoq, and maybe Barras's work.

  The 20,000 - 30,000 LoC mechanization are all about small languages
  with pretty much the same features as our language except for your
  second and third bullet point. martin-lof a la coq, Abel's work, and
  nbe in coq aren't that richer in feature otherwise. None includes
  cumulativity (they only have one predicative universe)

  The 400,000 NuPRL in Coq probably falls into a different cateogry
  because they are trying to mechanize a full practical language
}

}

@@

Finally, despite our earlier claim that our metatheoretic properties
are not as strong as some of the related work, we can strengthen our
results through syntactic means. For example, the
normalize-and-compare strategy does not induce an efficient algorithm
for type conversion, like the ones from \citet{decagda} and
\citet{martin-lof-a-la-coq}. However, with the standard syntactic
techniques from \citet{takahashi-parallel-reduction,
factorization-essentially}, we can prove that leftmost-outermost
reduction is a normalizing reduction strategy, from which we can
separately show the correctness a more efficient algorithm that
reduces the two terms to weak head normal form before recursively
comparing their subcomponents. We believe factoring such properties
out of a logical relation is valuable, as it helps us identify the
part of our proof that requires extra strength from the metatheory.
\scw{Why don't we just use leftmost-outermost reduction in the first place?
Do we even need nondeterministic parallel reduction?}
\yl{The conversion uses full reduction. Nondeterministic reduction
  makes it harder to show that convertible types have the same
  meaning. Maybe it would require us to prove factorization in our
  development but it definitely simplifies the determinism proof
  (confluence is no longer required before the fundamental
  lemma). }
